---
title: "The Exponential Distribution and the CLT"
author: "Marco Totolo"
output: html_document
---

## Overview

This document will show how the distribution of averages of N exponential random variables follows the Central Limit Theorem. This means that as the number of sample means increases, the distribution asymptotically tends to a normal distribution with mean equal to the exponential mean and standard deviation equal to the exponential standard deviation divided by the square root of N.

##Simulations

Let's first build an array of means of random exponential variables.

As specified in the assignment, I will take 1000 means of 40 random exponential variables with lambda 0.2.

```{r}
set.seed(4)
lambda<-0.2
N<-40
nosim<-1000
expMeans <- NULL
for (i in 1 : nosim) {
      expMeans = c(expMeans, mean(rexp(N,lambda)))
}
```

##Sample Mean versus Theoretical Mean

The sample mean is 
```{r}
mean(expMeans)
```
which is very close to the the CLT theoretical value, the mean of the distribution of the averages of exponential variables,  1/lambda
```{r}
1/lambda
```
To show the CLT in action let's repeat the 1000 simulations with 3 different values of N. The histogram will be more concentrated near the theoretical value as N increases. 
```{r}
library(ggplot2)
set.seed(8)
lambda<-0.2
Nvector<-c(10,40,100)
nosim<-1000
means<-NULL
for (i in Nvector) {
      N<-i
      for (j in 1 : nosim) {
            means <- c(means, mean(rexp(N,lambda)))
      }
}
meansData <- data.frame(means=means,N=rep(Nvector,each=nosim))
g<- ggplot(meansData,aes(means))
g+facet_grid(.~N)+
      geom_histogram(bins=30,color="black",fill="red",alpha=0.5)+
      geom_vline(xintercept = 1/lambda,size=1)
```

##Sample Variance versus Theoretical Variance

The sample variance is 
```{r}
var(expMeans)
```
which is very close to the the CLT theoretical value, the mean of the distribution of the averages of exponential variables,  1/(lambda^2*N)
```{r}
1/(lambda^2*N)
```
To show the CLT in action let's repeat the 1000 simulations with 3 different values of N. The histogram will be more concentrated near the theoretical value as N increases. 
```{r}
library(ggplot2)
set.seed(16)
lambda<-0.2
Nvector<-c(10,40,100)
nosim<-1000
variances<-NULL
for (i in Nvector) {
      N<-i
      for (j in 1 : nosim) {
            variances <- c(variances, mean(rexp(N,lambda)))
      }
}
varData <- data.frame(variances=variances,N=rep(Nvector,each=nosim))
g<- ggplot(varData,aes(variances))
g+facet_grid(.~N)+
      geom_histogram(bins=30,color="black",fill="red",alpha=0.5)+
      geom_vline(xintercept = 1/(lambda^2*N),size=1)
```

##Distribution

```{r}
lambda<-0.2
N<-40
nosim<-1000
theorMean<-1/lambda
theorSd<-1/(lambda*sqrt(N))
expMeans <- NULL
for (i in 1 : nosim) {
      expMeans = c(expMeans, mean(rexp(N,lambda)))
}
sampleMean <-mean(expMeans)
sampleSd<-sd(expMeans)
      
```


g+geom_histogram(bins=20,color="black",fill="red",alpha=0.5)+geom_vline(xintercept = 5,size=1.5)